---
title: "Lab 07 — Model Selection and Validation"
author: "EAES 480 — Modern Statistics in Earth & Environmental Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: inline
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)
library(broom)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Overview

This lab extends your linear modeling work from **fitting and interpreting regression models** to **comparing models**.

In Part I, you will evaluate model fit using:

- **sum of squares**
- **R-squared**
- **MSE**
- **RMSE**
- and **residual diagnostics**

You will work with the **US-AMS simplified AmeriFlux dataset** again and choose your own response and predictor variables.

Part II will later extend this work to **train-test validation splits**.

---

# Learning goals

By the end of this lab, you should be able to:

- Fit and compare multiple linear regression models
- Use `tidy()`, `glance()`, and `augment()` to summarize models
- Compute **SST**, **SSE**, and **R-squared**
- Compute **MSE** and **RMSE**
- Compare models using both statistical summaries and plots
- Explain why model selection requires more than one metric
- Produce a fully reproducible R Markdown analysis

---

# Data

## Load and inspect

```{r load_data, echo=TRUE, eval=FALSE}
# GOAL: Read the dataset and keep 2023 observations.

df <- read_csv("data/us-ams-simple.csv", na = c("-9999")) %>%
  clean_names() %>%
  filter(year_local == 2023)

glimpse(df)
```

## Derive time columns if needed

```{r derive_time, echo=TRUE, eval=FALSE}
# GOAL: Create date / month columns if they do not already exist.

df <- df %>%
  mutate(
    date = as.Date(floor(doy) - 1, origin = paste0(year_local, "-01-01")),
    month = month(date, label = TRUE, abbr = TRUE)
  )
```

---

# Part I — Model selection and assessment

## 1) Choose a response variable and candidate predictors

Pick:
- **one numeric response variable**
- **two candidate predictor variables**

Possible examples:
- response: `le`, `fc`, `gpp`
- predictors: `ta`, `swc`, `vpd`, `rg`

```{r choose_vars, echo=TRUE, eval=FALSE}
response_var <- "gpp"
pred1 <- "ta"
pred2 <- "swc"

# CHECK: do these variables exist?
c(response_var, pred1, pred2) %in% names(df)
```

> Which variables did you choose, and why do you think they may be related?

The variables I chose were `gpp` for my response variable and `ta` and `swc` as my predictor variables. I chose air temperature and soil water content as predictor variables of gross primary productivity since the productivity of plants and their carbon sequestration can be correlated to how much water they recieve from the ground and temperatures they can survive in. 

---

## 2) Create a clean analysis dataset

Keep only rows with non-missing values for your chosen variables.

```{r make_analysis_df, echo=TRUE, eval=FALSE}
# GOAL: Create an analysis data frame using only complete cases.

df_mod <- df %>%
  select(gpp, ta, swc) %>%
  drop_na()

glimpse(df_mod)
```

---

## 3) Fit three models

Fit:
- **Model 1:** response ~ pred1
- **Model 2:** response ~ pred2
- **Model 3:** response ~ pred1 + pred2

```{r fit_models, echo=TRUE, eval=FALSE}
m1 <- lm(gpp ~ ta, data = df_mod)
m2 <- lm(gpp ~ swc, data = df_mod)
m3 <- lm(gpp ~ ta + swc, data = df_mod)
```

---

## 4) View coefficients with `tidy()`

```{r tidy_models, echo=TRUE, eval=FALSE}
tidy(m1)
tidy(m2)
tidy(m3)
```

> Which coefficients are positive? Which are negative? Do they make scientific sense?

The positive coefficient in the data is `ta`, with a coefficient of around 0.546 and `swc` is negative with a coefficient of around -0.0657. I think the `ta` coefficient makes sense since as conditions become warmer, plants typically become more productive (to an extent). However, I do not think the `swc` coefficient makes sense since plants would be more productive with more available moisture.  

---

## 5) Visualize at least two one-predictor models

Make scatterplots with fitted linear trends for Model 1 and Model 2.

```{r plot_m1, echo=TRUE, eval=FALSE}
ggplot(df_mod, aes(x = .data[[pred1]], y = .data[[response_var]])) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic(base_size = 18) +
  labs(
    x = pred1,
    y = response_var,
    title = "Model 1"
  )
```

```{r plot_m2, echo=TRUE, eval=FALSE}
ggplot(df_mod, aes(x = .data[[pred2]], y = .data[[response_var]])) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic(base_size = 18) +
  labs(
    x = pred2,
    y = response_var,
    title = "Model 2"
  )
```

> Visually, which model seems stronger? Why?

The model that appears to be stronger is Model 1, since the correlation between the two variables appears to be stronger. Looking at the distribution in Model 2, the points are a lot more uniform and the correlation is more flatter, while still being negative. In Model 1, the points are less uniformly distributed and the slope is steeper, indicating a stronger relationship.

---

## 6) Get fitted values and residuals with `augment()`

```{r augment_models, echo=TRUE, eval=FALSE}
aug_m1 <- augment(m1, data = df_mod)
aug_m2 <- augment(m2, data = df_mod)
aug_m3 <- augment(m3, data = df_mod)

aug_m1 %>% select(.fitted, .resid) %>% head()
```

---

## 7) Compute sum of squares manually for Model 1

Recall:

- SST = total sum of squares
- SSE = sum of squared errors
- SSR = regression sum of squares

And:

$SST = SSR + SSE$

```{r ss_model1, echo=TRUE, eval=FALSE}
y <- df_mod[[response_var]]
ybar <- mean(y)

sst <- sum((y - ybar)^2)
sse_m1 <- sum(aug_m1$.resid^2)
ssr_m1 <- sst - sse_m1

tibble(
  sst = sst,
  sse_m1 = sse_m1,
  ssr_m1 = ssr_m1
)
```

> In words, what do SST and SSE represent?

SST represents the total amount of variability in the dataset in regards to the response variable. SSE represents any variation that is unexplained. 

---

## 8) Compute R-squared manually for Model 1

Use:

$R^2 = 1 - SSE / SST$

```{r r2_manual, echo=TRUE, eval=FALSE}
r2_m1_manual <- 1 - sse_m1 / sst

r2_m1_manual
```

## 9) Compare to `glance()`

```{r glance_model1, echo=TRUE, eval=FALSE}
glance(m1)
```

> Does your manual R-squared match the value from `glance()`?

My manual value does match the value from `glance()`. 

---

## 10) Compute MSE and RMSE for Model 1

Use:

- $MSE = mean(e_i^2)$
- $RMSE = sqrt(MSE)$

```{r mse_rmse_m1, echo=TRUE, eval=FALSE}
mse_m1 <- mean(aug_m1$.resid^2)
rmse_m1 <- sqrt(mse_m1)

tibble(
  mse_m1 = mse_m1,
  rmse_m1 = rmse_m1
)
```

> What does RMSE mean in the units of your response variable?

RMSE represents the difference between predicted values and the actual observed data points. Since it is generally low, that means the model is relatively accurate.  

---

## 11) Compare all three models using `glance()`

```{r compare_glance, echo=TRUE, eval=FALSE}
bind_rows(
  glance(m1) %>% mutate(model = "m1"),
  glance(m2) %>% mutate(model = "m2"),
  glance(m3) %>% mutate(model = "m3")
) %>%
  select(model, r.squared, adj.r.squared, sigma, AIC, BIC)
```

> Which model has the highest R-squared? Is that enough to call it the best model?

The model with the highest R-squared value is Model 3, but that is not enough information to call it the best model. It just means that there is the least variance in the data points between the response and predictor variable. 

---

## 12) Compare all three models using RMSE

```{r compare_rmse, echo=TRUE, eval=FALSE}
rmse_tbl <- bind_rows(
  augment(m1, data = df_mod) %>%
    summarise(rmse = sqrt(mean(.resid^2))) %>%
    mutate(model = "m1"),
  augment(m2, data = df_mod) %>%
    summarise(rmse = sqrt(mean(.resid^2))) %>%
    mutate(model = "m2"),
  augment(m3, data = df_mod) %>%
    summarise(rmse = sqrt(mean(.resid^2))) %>%
    mutate(model = "m3")
)

rmse_tbl
```

> Which model has the smallest RMSE? Does that agree with the R-squared ranking?

The model with the smallest RMSE value is Model 3, which does  agree with the R-squared ranking, as Model 3 had the largest R-squared value. 

---

## 13) Compare residual plots

Make residual vs fitted plots for at least two models.

```{r resid_plot_m1, echo=TRUE, eval=FALSE}
ggplot(aug_m1, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic(base_size = 18) +
  labs(
    x = "Fitted values",
    y = "Residuals",
    title = "Residuals: Model 1"
  )
```

```{r resid_plot_m3, echo=TRUE, eval=FALSE}
ggplot(aug_m3, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic(base_size = 18) +
  labs(
    x = "Fitted values",
    y = "Residuals",
    title = "Residuals: Model 3"
  )
```

> Do the residual plots suggest any problems (curvature, changing variance, outliers)? Which model looks more appropriate?

The residual plots suggest --------------

---

## 14) Make a model selection decision

Choose the model you think is best supported by the evidence.

Fill in:

> My preferred model is `Model 3`

> I based this decision on:
> - `RMSE value`
> - `R-Squared value`
> - `Residuals Plot`

**Prompt (5–7 sentences):** Explain why you chose this model. Use at least two quantitative criteria (such as R-squared, RMSE, or sigma) and one visual criterion (such as the residual plot or scatterplot). Also explain why a different model was less convincing.

> I chose Model 3 as my preferred model because ---------------. 

---


# Part II — Train/Test Splits and Cross-Validation

In this section you will extend model selection toward **prediction**.

You will:

- choose **two new predictors** (not `ta` or `swc`) to explain `le`
- hold out a **test set**
- use **4-fold cross-validation on training data**
- choose between **two candidate models**
- refit the chosen model on all training data
- evaluate it once on the held-out test data

The key idea is:

> Use training data to build models, validation folds to choose among them, and test data only once for final evaluation.

---

## 15) Choose two new predictors for `le`

Choose **two predictors that are NOT `ta` or `swc`**.

Possible examples depend on your file, but might include things like:
- `vpd`
- `rg`
- `fc`
- `gpp`
- `h`
- `rh`

```{r choose_new_predictors, echo=TRUE, eval=FALSE}
response_var <- "le"
pred1 <- "fc"
pred2 <- "gpp"

# CHECK: do these exist?
c(response_var, pred1, pred2) %in% names(df)
```

> Why did you choose these two predictors?

I chose these two predictors because both carbon in the atmosphere and gross primary productivity can influence the amount of latent heat observed in the atmosphere, and flux carbon and gpp are also closely linked variables.

---

## 16) Create a clean modeling dataset

Keep only complete cases for the response and the two predictors.

```{r make_prediction_df, echo=TRUE, eval=FALSE}
df_pred <- df %>%
  select(le, gpp, fc) %>%
  drop_na()

glimpse(df_pred)
```

> How many observations remain after removing missing values?

The same number of observations, which is 16,973, remained after removing the missing values. 

---

## 17) Shuffle rows and create a train/test split

Use 20% of the data as a final **test set**.

```{r split_train_test, echo=TRUE, eval=FALSE}
set.seed(41)

df_shuf <- df_pred %>%
  slice_sample(prop = 1)

n_total <- nrow(df_shuf)
n_test <- round(0.20 * n_total)

test_data <- df_shuf %>% slice(1:n_test)
train_data <- df_shuf %>% slice((n_test + 1):n_total)

tibble(
  dataset = c("train", "test"),
  n = c(nrow(train_data), nrow(test_data))
)
```

> Why do we keep the test set separate from model selection?

You keep the test set separate from the model selection because the isolation ensures the model is unbiased towards the data. 

---

## 18) Define two candidate models

Use:

- **Model 1:** `le ~ pred1`
- **Model 2:** `le ~ pred1 + pred2`

Write them out in words below:

> Model 1: `Linear regression model using flux carbon to predict latent heat`

> Model 2: `Linear regression model using flux carbon and gross primary productivity to predict latent heat`

---

## 19) Create 4 folds on the training data

```{r make_cv_folds, echo=TRUE, eval=FALSE}
set.seed(41)

train_folds <- train_data %>%
  mutate(fold = rep(1:4, length.out = n()))

train_folds %>% count(fold)
```

> Why do we create folds only on the training data, and not using the test data?

We create folds only on the training data because it prevents overfitting the data, or making the model capture random fluctuations in the data rather than the important patterns. 

---

## 20) Run 4-fold cross-validation

For each fold:

- train on 3 folds
- validate on 1 held-out fold
- compute RMSE

```{r cv_loop, echo=TRUE, eval=FALSE}
cv_results <- map_dfr(1:4, function(k) {

  cv_train <- train_folds %>% filter(fold != k)
  cv_valid <- train_folds %>% filter(fold == k)

  m1 <- lm(le ~ fc, data = cv_train)
  m2 <- lm(le ~ fc + gpp, data = cv_train)

  tibble(
    fold = k,
    model = c("m1", "m2"),
    rmse = c(
      sqrt(mean((cv_valid[[response_var]] - predict(m1, newdata = cv_valid))^2)),
      sqrt(mean((cv_valid[[response_var]] - predict(m2, newdata = cv_valid))^2))
    )
  )
})

cv_results
```

---

## 21) Summarize cross-validation performance

Compute the mean and SD of RMSE across folds.

```{r cv_summary, echo=TRUE, eval=FALSE}
cv_summary <- cv_results %>%
  group_by(model) %>%
  summarise(
    mean_rmse = mean(rmse),
    sd_rmse = sd(rmse),
    .groups = "drop"
  )

cv_summary
```

> Which model has the lower average validation RMSE?  
> What does the SD across folds tell you?

The model with the lower average RMSE is Model 2. The SD across folds tells me ------------------. 

---

## 22) Visualize fold-by-fold validation RMSE

```{r plot_cv_results, echo=TRUE, eval=FALSE}
ggplot(cv_results, aes(x = factor(fold), y = rmse, color = model, group = model)) +
  geom_point(size = 3) +
  geom_line() +
  theme_classic(base_size = 18) +
  labs(
    x = "Held-out validation fold",
    y = "Validation RMSE",
    title = "Cross-validation performance across folds"
  )
```

> Does one model win consistently across folds, or only sometimes?

Model 2 consistently wins across folds because it's values are consistently lower ----. 

---

## 23) Choose your preferred model

Fill in:

> My selected model is: `Model 2`

> I chose it because: `___`

---

## 24) Refit the chosen model on all training data

Use all of `train_data` now that you have selected the model.

```{r refit_final_model, echo=TRUE, eval=FALSE}
final_model <- lm(le ~ fc + gpp, data = train_data)

tidy(final_model)
```

---

## 25) Evaluate the final model on the test set

Use `predict()` and compute test RMSE.

```{r test_evaluation, echo=TRUE, eval=FALSE}
test_pred <- predict(final_model, newdata = test_data)

test_rmse <- sqrt(mean((test_data[[response_var]] - test_pred)^2))

test_rmse
```

> How does the test RMSE compare with the cross-validation RMSE?

---

## 26) Final interpretation

Answer in **5–7 sentences**:

- Why did you not use the test data during model selection?
- How did cross-validation help you choose between models?
- Did the chosen model generalize well to test data?
- What would it mean if test RMSE were much worse than validation RMSE?

> *Write your answer here.*

---

# Submission

- Knit your `.Rmd` to HTML.
- Commit and push both the `.Rmd` and knitted `.html` to your GitHub repo.
- Submit the GitHub repository link on Canvas.
